---
title: "ACCC-CRC WES Analysis"
number-sections: true
include-after-body: "footer.html"
format: 
 html:
  code-fold: true
  code-summary: "Show the code"
author: "Efstathios-Iason Vlachavas"
date: last-modified
date-format: YYYY-MM-DD
toc: true
theme:
 dark: darkly
 light: flatly
editor: visual
warning: false
---

# ACCC-CRC dataset WES analysis overview

### Reproduce the analysis of the rnaseq data from the main ACCC-CRC dataset for the manuscript.

Initially start by loading the necessary R packages, along with importing the clinical information.

```{r}

library(ComplexHeatmap)
library(InteractiveComplexHeatmap)
library(tidyverse)
library(circlize)
library(ggthemes)
library(here)

```

### Create an oncoprint with selected IntOGen CRC driver genes (**Figure 1**)

```{r,warning=FALSE, out.width="200px"}

dr_here()

# load the matrix with the 54 altered driver genes-
dt <- read_csv(here("Data","Input_Oncoplot_ACCC_CRC.csv"))

dt2 <- dt |> column_to_rownames("hugo") # here
# na= character()
mat <- dt2
mat[is.na(mat)] = ""
mat = as.matrix(mat)

# approach of total alteration frequency counting of the selected 54 genes
# keep caution for the counting process to remove any entries in the value column with NA !!

dat_count <- dt |> pivot_longer(cols=-hugo) |> filter(!is.na(value)) |> 
  distinct(name, hugo) |> group_by(hugo) |> count(hugo, name, sort=T) |> 
  count(hugo, sort=T)

pdat_updated <- read_tsv(here("Data",
"CRC.ACCC.Clinical.Total.Merged.MSI.Mut.Info.RNASeq.tsv"))

# here; just on this case we remove *manually* the CRC_ACCC_02 sample in order to project the results for the WES data {as we have collectively all clinical info for all profiled samples from both omic layers}

pdat.ord <- pdat_updated[order(pdat_updated$MSI_status, decreasing = TRUE),] |> 
select(Sample_Name,MSI_status) |> 
filter(!Sample_Name=="CRC_ACCC_02") |>
column_to_rownames("Sample_Name") 

s.ord <- as.character(rownames(pdat.ord))

col_2 =c("missense" = "#008000", "stop_gained"="black", "splice_site"="brown",
         "Gain"="red","Loss"="blue","frameshift_insert"="yellow","LOH"="#00FFFF",
         "frameshift_del"="orange","inframe_deletion"="khaki2",
         "start_lost"="purple")

#########################################################################################
# implement the following scheme to map alterations to distinct colors/rectangle combos!!
alter_fun_2 = list(
  background = function(x, y, w, h) {
    grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
              gp = gpar(fill = "#CCCCCC", col = NA))
  },
  # black
  stop_gained = function(x, y, w, h) {
    grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
              gp = gpar(fill = col_2["stop_gained"], col = NA))
  },
  
  # purple
  start_lost = function(x, y, w, h) {
    grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
              gp = gpar(fill = col_2["start_lost"], col = NA))
  },
  
  # brown
  splice_site = function(x, y, w, h) {
    grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), 
              gp = gpar(fill = col_2["splice_site"], col = NA))
  },
  
  # yellow
  frameshift_insert = function(x, y, w, h) {
    grid.rect(x, y, w-unit(0.5, "mm"), h*0.45, 
              gp = gpar(fill = col_2["frameshift_insert"], col = NA))
  },
  
  # cyan
  LOH = function(x, y, w, h) {
    grid.rect(x, y, w-unit(0.5, "mm"), h*0.5, 
              gp = gpar(fill = col_2["LOH"], col = NA))
  },
  
  # orange
  frameshift_del = function(x, y, w, h) {
    grid.rect(x, y, w-unit(0.5, "mm"), h*0.45, 
              gp = gpar(fill = col_2["frameshift_del"], col = NA))
  },
  
  # khaki2
  inframe_deletion = function(x, y, w, h) {
    grid.rect(x, y, w-unit(0.5, "mm"), h*0.4, 
              gp = gpar(fill = col_2["inframe_deletion"], col = NA))
  },
  
  # green
  missense = function(x, y, w, h) {
    grid.rect(x, y, w*0.7, h*0.8, 
              gp = gpar(fill = col_2["missense"], col = NA))
  }, 
  # red
  Gain = function(x, y, w, h) {
    grid.rect(x, y, w-unit(0.5, "mm"), h*0.5, 
              gp = gpar(fill = col_2["Gain"], col = NA))
  }, 
  # blue
  Loss = function(x, y, w, h) {
    grid.rect(x, y, w-unit(0.5, "mm"), h*0.5, 
              gp = gpar(fill = col_2["Loss"], col = NA)) 
    
  }
)

column_title = "OncoPrint {ACCC CRC cohort IntoGen: 54 Common Scored Genes}"
##############################################################################

# also load the estimated mutational spectrum from prior analyses; also check the related Zenodo material

mut.spect.final <- read_tsv(here("Data",
"MutSpec_TypeOccurence_ACCC_FinalSamples_Oncoplot.tsv")) |> 
column_to_rownames("ACCC_sample_ID") |> dplyr::select(-c(7,8))

mutSpect.scaled = t(apply(mut.spect.final, 1, function(x) x/sum(x)))
mutSpect.ord <- mutSpect.scaled[rownames(pdat.ord),]

# set the complete heatmap annotation object

ha <- HeatmapAnnotation(cbar = anno_oncoprint_barplot(),
      MSI_Status_RNASeq = ifelse(pdat.ord[[1]] == 1, "MSI","MSS"),
      col=list(MSI_Status_RNASeq = structure(names = c("MSI", "MSS"), c("black", "red"))),
      Mutational_Spectrum = anno_barplot(mutSpect.ord, 
      gp = gpar(fill = c(5,1,2,8,3,6)), bar_width = 1, height = unit(3, "cm")))

lgd = Legend(title = "Mutational_Spectrum", labels = c("C>A","C>G","C>T","T>A","T>C","T>G"), 
             legend_gp = gpar(fill = c(5,1,2,8,3,6)))

mat2 <- mat[dat_count$hugo,rownames(pdat.ord)]

# set the final oncoplot

ht = oncoPrint(mat2,alter_fun = alter_fun_2, col = col_2, top_annotation = ha, 
               row_order = dat_count$hugo, column_names_gp = gpar(fontsize = 7),
               column_title = column_title, column_order=s.ord,
               remove_empty_columns = FALSE)

draw(ht, heatmap_legend_list = lgd)

```

<button type="button" class="btn btn-primary" data-bs-toggle="modal" data-bs-target="#exampleModal">

See in Full Screen

</button>

::: {#exampleModal .modal .fade tabindex="-1"}
::: {.modal-dialog .modal-dialog-centered .p-8 style="max-width: 90%"}
::: modal-content
```{r, echo=FALSE, warning=FALSE, out.width="100%", fig.width=12, fig.height=10}
draw(ht, heatmap_legend_list = lgd)
```
:::
:::
:::

### Code for Supplementary plot regarding relative contribution of COSMIC Mutational Signatures {**Supplementary Figure 1**}

```{r,fig.width=9}

sig.contrib <- read_csv(here("Data","Signature_contributions_ACCC.csv"))
sig.contrib <- sig.contrib |> select(-Proposed_Etiology) |> pivot_longer(cols=-Signature)

# use the imported clinical information object from above chunk
clin.dat <- pdat_updated |> 
select(Sample_Name, MSI_status)

msi_accc <- as.character(clin.dat$MSI_status)
msi_accc_upd <- recode(msi_accc, "0" = "MSS", "1" = "MSI")

clin.dat$MSI_status <- msi_accc_upd

xx.1 <- left_join(x=sig.contrib, y=clin.dat, by=c("name"="Sample_Name"))
xx.1 <- xx.1 |> arrange(MSI_status)

## how to *manually* group respectively the assigned COSMIC SBS signatures (https://cancer.sanger.ac.uk/signatures/sbs/)::

## SBS6, SBS20, SBS21, SBS26 and SBS15 -> MSI -> one color
## SBS1 + SBS5 -> keep them as separate signatures but use similar colors to show common-partial-etiologies -> clock-like association (mainly because the etiologies are different although both associate with age)
## SBS10 a + b -> Group together
## SBS93 + SBS94 since the etiology is unknown (and probably different) -> separate but similar C;
## check this-below is the most optimal with ordering and color specification:

xx.2 <- xx.1 |> distinct(name,MSI_status)

a <- a <- ifelse(xx.2$MSI_status == "MSS", "black", "red") # this to also separate-order samples with MSI/MSS samples

# This might differ slightly from the original supplementary figure, as here to enhance differences and improve explainability, samples are also colored + ordered based on MSI status;

  ggplot(xx.1, aes(fill=Signature, y=value, x=factor(name,
  levels=c("CRC_ACCC_09","CRC_ACCC_15","CRC_ACCC_17","CRC_ACCC_21","CRC_ACCC_24","CRC_ACCC_01",
           "CRC_ACCC_03","CRC_ACCC_04","CRC_ACCC_05","CRC_ACCC_06","CRC_ACCC_07","CRC_ACCC_08",
           "CRC_ACCC_11","CRC_ACCC_12","CRC_ACCC_13","CRC_ACCC_14","CRC_ACCC_16","CRC_ACCC_18",
           "CRC_ACCC_19","CRC_ACCC_20","CRC_ACCC_22","CRC_ACCC_23","CRC_ACCC_25",
           "CRC_ACCC_26","CRC_ACCC_27","CRC_ACCC_28","CRC_ACCC_29")))) +
  geom_bar(position='stack', stat='identity') + 
  theme_wsj() + theme(axis.text.x = element_text(angle = 90, hjust = 1, colour = a)) + 
  ggtitle("COSMIC SBS Signatures") +
  scale_fill_manual(values = c("darkseagreen2", "gold4","gold4","bisque",
                               "gold2", "cyan", "coral", "gold2", "gold2",
                               "#bdb76b", "gold2", "aquamarine4", "chartreuse4",
                               "brown","darkgray", "darkseagreen", "darkgray",
                               "darkgray", "gold2", "darkviolet","darkorchid1"))

```

#### Finally we proceed with presenting a *condensed* view of the *scored + actionable* variants in the ACCC-CRC cohort at the patient resolution, via an *enhanced* oncoprint view; This is also a major part of the final **Figure 5** panel;

```{r}
#| fig-align: center
#| fig-width: 14
#| fig-height: 10
#| fig.cap: "Oncoplot with actionable variants (SVRACAS score ≥0.5) and annotated in the context of sensitivity (OncoKB_Sens 1-4) or resistance (OncoKB_Res R1-2) to a targeted therapy (https://www.oncokb.org/levels) in any tumor entity."
#| message: false

# Based on the main text and relative section, here we sought to define the actionable alterations per patient cluster, and we ultimately assign them also to available MASTER biomarker baskets: 

out.CMS.ACCC <- read_tsv(here("Data","Out_RES_DT_CMScaller_ACCC_28samples.tsv"))
out.CMS.ACCC.sel <- out.CMS.ACCC %>% dplyr::select(ACCC_Sample_ID, prediction)

TMB_info <- read_tsv(here("Data","ACCC_CRC_TMB_INFO.txt"))

cluster_info <- read_tsv(here("Data","HCPC_FactoMineR_RES_ACCC_CRC.tsv"))

cluster_info_merge <- inner_join(cluster_info, out.CMS.ACCC.sel, by=c("Sample_ID"="ACCC_Sample_ID"))

cluster_info_merge_TMB <- inner_join(cluster_info_merge, TMB_info, by=c("Sample_ID"="CRC_ACCC_Patient_ID"))

cluster_info_sel <- cluster_info_merge_TMB |> select(Sample_ID, MSI_status, Condition, clust, prediction, `Mutational_Burden (Mutations/Mb)`)

cluster_info_ord <- cluster_info_sel[order(cluster_info_sel$clust, 
                    cluster_info_sel$MSI_status, cluster_info_sel$Condition),]

# extra line to remove here patient ACCC_CRC_02 to have the finalized complex Heatmap {as in the overall clinical information, we kept ACCC_CRC_02, due to the RNASeq profiling data presence}

cluster_info_ord <- cluster_info_ord |> dplyr::filter(!Sample_ID=="CRC_ACCC_02")

################################################################################
# import MASTER baskets prior-knowledge information {for further information check also here: https://www.nct-heidelberg.de/en/the-nct/newsroom/news/details/dkfz-nct-dktk-master-program-handbook-on-the-use-of-genomic-data-in-precision-oncology.html}

# the following is available supplementary file from the following publication: https://doi.org/10.1158/2159-8290.cd-21-0126
master.baskets <- read_tsv(here("Data","MASTER.Baskets.biomarkers.SuppTableS5.2022.txt"))

# filter for profiled genes with *SNV/indel* hits
master.baskets.sel <- master.baskets |> dplyr::filter(Alteration=="SNV/indel") |> 
dplyr::select(Biomarker, Biomarker_basket) |> drop_na()

master.bim.baskets.unique <- master.baskets.sel |> distinct(Biomarker, Biomarker_basket, .keep_all = TRUE)

# load both MSI + MSS actionable files to be merged downstream:

MSS.OnkoKB.scored.actionable <- read_tsv(here("Data",
"ACCC.CRC.Scored.MSS.OnkoKB.Merged.Paper.StrictFilter.Resistance.Sensitive.tsv"))

# similarly for the MSI samples
MSI.OnkoKB.scored.actionable <- read_tsv(here("Data",                                          "ACCC.CRC.Scored.MSI.StrictFilter.Resistance.Sensitive.OnkoKB.Merged.Paper.tsv"))

# to combine all alterations into 1 merged table
total.MSI.MSS.actionable <- full_join(MSS.OnkoKB.scored.actionable,MSI.OnkoKB.scored.actionable)

# Then create the individual heatmap + row annotations as a *complex* visualization means for charting the precision oncology landscape of the ACCC-CRC cohort

# Here for downstream total heatmap analysis
master.combo.dat <- left_join(total.MSI.MSS.actionable, master.bim.baskets.unique, by=c("hugo"="Biomarker"))

## Now focus and try to create the additional layers of actionable mutations from OncoKB;
## For more information please check here: https://www.oncokb.org/

# first supplementary heatmap for Sensitivity hits
Annot_Sensitive <- total.MSI.MSS.actionable |> select(Protein_alt, hugo, samples, `Highest Sensitive Level`) |> mutate(Feature=str_c(hugo,Protein_alt, sep = "_")) |>
distinct(Feature, `Highest Sensitive Level`) |> relocate(Feature, .before=`Highest Sensitive Level`) |> deframe()

# second-like supplementary heatmap for Resistance hits
Annot_Resistance <- total.MSI.MSS.actionable |> select(Protein_alt, hugo, samples, `Highest Resistance Level`) |> mutate(Feature=str_c(hugo,Protein_alt, sep = "_")) |>
distinct(Feature, `Highest Resistance Level`) |> relocate(Feature, .before=`Highest Resistance Level`) |> deframe()

# main heatmap body
main_onko_dat <- total.MSI.MSS.actionable |> select(samples, so, Protein_alt, hugo) |> 
mutate(Feature=str_c(hugo,Protein_alt, sep = "_")) |> select(Feature, samples, so) |>
pivot_wider(names_from = "samples", values_from="so",values_fill=NA) |> column_to_rownames("Feature")

# the heatmap with the MASTER baskets
MASTER_baskets_dat <- master.combo.dat |> select(Protein_alt, hugo, Biomarker_basket) |> 
mutate(Feature=str_c(hugo,Protein_alt, sep = "_")) |> mutate(Biomarker_basket=as_factor(Biomarker_basket),value=1) |> select(Feature, Biomarker_basket,value) |> distinct(Feature, Biomarker_basket, .keep_all = TRUE) |> pivot_wider(names_from = "Biomarker_basket", values_from="value", values_fill=0) |> column_to_rownames("Feature")

##########################################################################################
# extra step for the MASTER baskets-patients matching:

out_MASTER_patients <- master.combo.dat |> dplyr::select(samples, Biomarker_basket) |> 
mutate(Samples_group=as_factor(samples),value=1) |> dplyr::select(Samples_group,Biomarker_basket,value) |> distinct(Samples_group,Biomarker_basket, .keep_all = TRUE) |> pivot_wider(names_from = "Samples_group", values_from="value", values_fill=0) |> column_to_rownames("Biomarker_basket")

##########################################################################################
# just small modification for the main heatmap body, necessary for visualization purposes
dat <- main_onko_dat
dat[is.na(dat)] = ""
mat = as.matrix(dat)

# here also keep and subset the total clinical dat with only the samples that were identified with at least one actionable alteration

final_snames <- colnames(mat)

pdat_final <- cluster_info_ord |> dplyr::filter(Sample_ID%in%final_snames) |> column_to_rownames("Sample_ID")

s.ord <- rownames(pdat_final)

col_2 =c("missense_variant" = "#008000", "stop_gained"="black", "splice_site_variant"="brown",
         "frameshift_elongation"="yellow","frameshift_truncation"="orange",
         "start_lost"="purple")

column_title = "OncoPrint {ACCC-CRC cohort: 17 patients}"

cl_head = "ACCC-CRC clusters: MSI/MSS OncoKB scored actionable mutations"

alter_fun_3 = list(
    background = alter_graphic("rect", fill = "#CCCCCC"),   
    stop_gained = alter_graphic("rect", fill = col_2["stop_gained"]),
    splice_site_variant = alter_graphic("rect", height = 0.6, fill = col_2["splice_site_variant"]),
    frameshift_elongation = alter_graphic("rect", fill = col_2["frameshift_elongation"]),
    frameshift_truncation = alter_graphic("rect", height = 0.33, 
    fill = col_2["frameshift_truncation"]),
    missense_variant = alter_graphic("rect", height = 0.33, fill = col_2["missense_variant"]),
    start_lost = alter_graphic("rect", height = 0.33, fill = col_2["start_lost"])
) 

ha_add = HeatmapAnnotation(
    Cluster_ID = pdat_final[[3]],
    MSI_status = ifelse(pdat_final[[1]] == 1, "MSI","MSS"),
    Mutation_group = pdat_final[[2]],
    CMS_subtype = pdat_final[[4]],
    TMB = ifelse(pdat_final[[5]] >= 10, "TMB-High", "TMB-Low"),
    border = TRUE,
    col = list(Mutation_group = structure(names = c("BRAF_mut", "KRAS_mut", 
    "GNAS_mut", "NRAS_mut", "RAS_RAF_GNAS_wt", "RAS_RAF_GNAS_mut"), 
    c("yellow", "red", "black", "green", "orange", "cyan")),
    MSI_status = structure(names = c("MSI","MSS"), c("red","black")),
    Cluster_ID = structure(names = c("1","2","3"),c("black","red","blue")),
    CMS_subtype = structure(names = c("CMS1","CMS2","CMS3","CMS4"),                                   c("orange","blue","purple","darkgreen")),
    TMB = structure(names=c("TMB-High", "TMB-Low"), c("magenta","yellow")))) 


col_letters = c("1" = "red", "2" = "purple", "3A" = "orange", "4"="pink")

col_letters_3 = c("1" = "black", "NA"="white")

############################################################################################
############# The extra MASTER-patients-baskets assignment heatmap #########################
############################################################################################

hmap_MASTER <- Heatmap(out_MASTER_patients[,s.ord],
                     name = "MB_enrichment",
                     col = c('0' = 'white', '1' = 'cornflowerblue'),
                     rect_gp = gpar(col = 'grey85'),
                     cluster_rows = TRUE,
                     show_row_dend = FALSE,
                     row_title = 'MASTER_baskets',
                     row_title_gp = gpar(fontsize = 11, fontface = 'bold'),
                     row_title_rot = 90,
                     row_title_side = "left",
                     show_row_names = TRUE,
                     row_names_gp = gpar(fontsize = 12),
                     row_names_side = 'left',
                     row_dend_width = unit(35, 'mm'),
                     cluster_columns = FALSE,
                     show_column_dend = FALSE,
                     column_title_gp = gpar(fontsize = 14, fontface = 'bold'),
                     column_title_rot = 0,
                     show_column_names = TRUE,
                     show_heatmap_legend = FALSE, 
                     clustering_distance_rows = 'euclidean',
                     clustering_method_rows = 'ward.D2')

### final part of creating the enhanced oncoprint figure:

### combine both *supplementary sensitivity/specificity* heatmaps from above into a merged row annotation

row_ha <- rowAnnotation(OncoKB_Sens=Annot_Sensitive[rownames(mat)], OncoKB_Res = Annot_Resistance[rownames(mat)], gp = gpar(fontsize = 18),
col = list(OncoKB_Sens = structure(names = c("1", "2", "3A", "4"), c("red","purple", "orange", "pink")), OncoKB_Res = structure(names=c("1"),c("black"))))

ht_main = oncoPrint(mat[,s.ord], alter_fun = alter_fun_3, col = col_2, top_annotation = ha_add,
column_title = cl_head, column_order=s.ord,show_column_names = TRUE, 
column_names_gp = gpar(fontsize = 7), row_names_gp = gpar(fontsize = 12), 
remove_empty_columns = FALSE, right_annotation = row_ha) # just removes the right barplot with the occurrence of alterations

ht_list = ht_main %v% hmap_MASTER
draw(ht_list)

# due to quatro/html display constraints, please check if this also works through the  InteractiveComplexHeatmap R package to create a separate interactive version:
# htShiny(ht_list)
# htShiny(ht_list, save = "test_ht_list")


```
